---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Import and Clean Data

## Selection Data

The original dataset was organized based on how the the pool was run with picks
in columns. We will want to make this much wider by having each golfer as a
column with a binary indicator of whether or not a given member of the pool
selected them.  We create some variable tests at the same time.

```{r}
data("entry_picks_anon")


wide_dat <- entry_picks_anon %>%
  pivot_longer(cols = -alt_name) %>%
  mutate(name = value,
         value = 1) %>%
  pivot_wider(names_from = name, values_from= value) %>%
  mutate_at(vars(-alt_name), ~ ifelse(is.na(.), 0, .)) 

report <- data_validation_report()
golfers <- entry_picks_anon %>% select(-alt_name) %>% unlist() %>% unique

test_picks <-
  all(wide_dat %>% mutate(tot_pick = rowSums(across(
    `Jon Rahm`:`Robert MacIntyre`
  ))) %>%
    pull(tot_pick) == 8)

validate(wide_dat, "Validate player selection data") %>%
  validate_if(description = "Verify everyone has 8 picks and nobody has any NA",
              all(test_picks),
              ) %>%
  validate_cols(description = "Test all values are 0 or 1",
                in_set(0, 1), -alt_name) %>%
  validate_if(description = "Test all 272 players are present",
              nrow(wide_dat) == 272) %>%
  validate_if(description = "Test all golders are present",
              all(golfers %in% colnames(wide_dat))) %>%
  add_results(report)


```

## Leaderboard Data

The leaderboard data has to be organized in order to allow for calculating the
daily ranks of each player. This data is pulled directly from the [ESPN
leaderboard]("https://www.espn.com/golf/leaderboard") using the `rvest` package. Code for html scraping can be found in
`/data-raw/leaderboard.R`. 

```{r}
data("leaderboard")

datatable(leaderboard,
          rownames = FALSE,
          colnames = c("Player", 
                       "Position", 
                       "Total Score", 
                       paste("Day", 1:4)))

```


The rules for the pool were not entirely clear with respect to handling players
that didn't make the second day cut, but it seemed to follow that they just kept
their original ranking from the end of day 2. This is a little different from
the official leaderboard, where DQs are dropped to the bottom. We use
`rowCumsums` to get the cumulative score for each player and then call a quick
custom ranking function to make sure ties are coded correctly. All golfers are
included in the first day and then those in the group that made the day 2 cut
are then separated out after that. Players who don't make the cut still have
their day 2 ranks carried forward.

```{r}
 
# Order of variables for later
order <- quo(c(player:tot, r1:r4, r1_rank, r2_rank, r3_rank, r4_rank))
  # Day 1:2
  pre_cut <- 
    leaderboard %>% dplyr::select(r1:r2) %>%
    as.matrix() %>%
    matrixStats::rowCumsums() %>% # Get cumulative score
    `colnames<-`(paste0("r",1:2)) %>%
    cbind(leaderboard %>% dplyr::select(-c(r1:r2))) %>%
    tibble::as_tibble() %>%
      ties(., var = r1) %>% # Score ties for day 1
      ties(., var = r2) # Score ties for day 2
    
    
  
  # Day 3 on limited to those who made the cut. Start from full leaderboard
  # to make sur eyou get the right cumulative scores
  made_cut <-  leaderboard %>%
    dplyr::filter(!is.na(pos))
  
  ranks_made <- made_cut %>%
     dplyr::select(r1,r2,r3,r4) %>%
     as.matrix() %>%
     matrixStats::rowCumsums() %>% # na.rm by default
     `colnames<-`(paste0("r",1:4)) %>%
     cbind(made_cut %>% dplyr::select(-c(r1,r2,r3,r4))) %>%
     tibble::as_tibble() %>%
     ties(., var = r3) %>%
     ties(., var = r4) %>%
    left_join(pre_cut %>% select(player, ends_with("rank"))) %>%
    select(!! order)
    
  
   ranks_cut <- pre_cut %>%
     dplyr::filter(is.na(pos)) %>%
    dplyr::mutate(r3_rank = r2_rank,
                  r4_rank = r2_rank) %>%
     select(!! order)

   
score_dat <-  rbind(ranks_made, ranks_cut)

# Add data tests
validate(score_dat, "Validate leaderboard data") %>%
validate_if(all(ranks_made$pos == ranks_made$r4_rank),
              description = "Verify ranks calculated by hand against the current position. If TRUE, assumes the other days are correct as well.") %>%
validate_if(description = "Test that highest score matches current score. Assume all the other days are correct if passes",all(ranks_made$r4 == ranks_made$tot)) %>%
  add_results(report)

```

## Calculate Daily Scores for Pool Participants

The final step of data prep is to combine the golfer daily scores with the
player choices in order to calculate the score of each day of the tournament.
This is calculated as the sum of the ranks of the players that each person
chose. For example if your 8 players are in positions `r paste(1:8, collapse = ", ")` 
then your score would be `r sum(1:8)`. A first step here however is to correct
some differences in the names between the pool and the ESPN leaderboard. These
are just some small spelling differences that are easier to change in leaderboard
data than the other.

```{r}

golfers[!golfers %in% score_dat$player]

score_dat <- score_dat %>%
  mutate(player = case_when(player == "Bryson DeChambeau" ~ "Bryson Dechambeau",
                            player == "Charles Osborne (a)" ~ "Ollie Osborne",
                            player == "Tyler Strafaci (a)" ~ "Tyler Strafaci",
                            TRUE ~ player))
```


The final step in data prep is to calculate the daily scores for everyone in the
pool. This is done by looping over rows in the pick data frame and and then
for each user calculate the sum of their player ranks for each day by matching
them to the score data. We test these against the ground-truth official scores
for a few key users and add those results to the validation report. There are
plenty of code inefficiencies here (row-wise operations are pretty slow) and
a better approach would be to also scrape the final results and confirm they
all match instead of just looking at this group. This will have to do for now
though!

```{r}
pick_dat <- entry_picks_anon %>% select(-alt_name)

final_scores <- apply(pick_dat, 1, function(x) {

  # For each set of picks iterate over days
  
  map(paste0("r",1:4,"_rank"), ~ {
    day <- .
   rank_day <- score_dat %>% select(player, all_of(day))

   sum(rank_day[match(x, rank_day$player), day])
 
    
  }) %>% do.call(cbind, .) %>% as_tibble %>% set_names(paste0("score_", 1:4))
    
}) %>% do.call(rbind, .)  %>% bind_cols(alt_name = entry_picks_anon$alt_name)

 
test_names <- c("Eastin, John" = 309, #ES1
                "Morlock, Geya" = 276, #ES2
                "Williams, Cameron" = 242, #ES3
                "Ross, Macie" = 252, #ES4
                "el-Afzal, Haazima" = 279 #ES5
                ) 

validate(final_scores, description = "Validate score calculation algorithm") %>%
  validate_if(description = "Validate final scores against official scores for ES staff", all(final_scores %>% filter(alt_name %in% all_of(names(test_names))) %>%
  pull(score_4) == test_names)) %>%
  add_results(report)


 
```

Before printing the validation report let's add one last detail that might be
interesting. We were provided picks within groups, which I (maybe naively) assumed was organized approximately by player ability. The first three picks 
were from group A, the next three from group B, and the final two from group C.

```{r}
group_a <- pick_dat[,1:3] %>% unlist %>% unique
group_b <- pick_dat[,4:6] %>% unlist %>% unique
group_c <- pick_dat[,7:8] %>% unlist %>% unique

group_dict <- tibble(group = c(rep("A", length(group_a)), 
                               rep("B", length(group_b)), 
                               rep("C", length(group_c))),
       player = c(group_a, group_b, group_c))

# Manually add 7 players from group C nobody chose
man <- tibble(
  group = "C",
  player = c(
    "Jim Herman",
    "Robert Streb",
    "Joe Long (a)",
    "Hudson Swafford",
    "Sandy Lyle",
    "Fred Couples",
    "Larry Mize"
  )
)

group_dict <- rbind(group_dict, man)

```


Finally we print the validation report

```{r}
print(report)
```

