---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Initial Data Exploration


## Golfer Performance

If we look at the leaderboard position of the full player field over time
we end up with something that is a little overwhelming. We can several ties
over time where ranks converge together and we can also see the cuts on day two 
from the straight lines from that point onwards. One thing I am a little surprised by in the beginning is that once you get out of the best few players
you actually get a lot of variability in ranks over days. This makes me regret
not taking a little more time to watch because I can see already that daily
scores in the pool would have had some potentially big swings.

```{r}
# Match style being used in the slides

rank_names <- c(r1_rank = "Day 1",
                r2_rank = "Day 2",
                r3_rank = "Day 3",
                r4_rank = "Day 4")

# Add indicator for players that are cut
score_dat <- score_dat %>%
  mutate(cut = ifelse(is.na(pos), TRUE, FALSE))

tx_def <-  function() {theme_xaringan(text_font_size = 10,
                 title_font_size = 10)
}


score_pdat <- score_dat %>%
  select(player, r1_rank:r4_rank, cut) %>%
  pivot_longer(-c(player, cut)) %>%
  mutate(name = factor(name, levels = names(rank_names), labels = rank_names)) %>%
  left_join(group_dict)

base_p <- score_pdat %>%
  ggplot(aes(x = name, y = value, group = player)) +
  labs(x = "Day of Tournament", y = "Leaderboard Position") +
  scale_y_reverse() + 
  geom_line() + 
  tx_def()
 
base_p + labs(title = "Leaderboard over time - All players")
  

```

I don't really watch a lot of golf so coming in I was expecting players that
were cut were generally would have had a bad tournament from the beginning.
But we can see that while this is broadly true there were actually a few players
who came out strong but had a bad day two. 

```{r}
base_p +
  gghighlight(cut == TRUE)
```

We can look a little closer at the development of the top ten players over time.
It looks like you get a mix of people who pretty much started and stayed there,
and others who had a bad first day before finishing strong. 
```{r}
base_p + 
  gghighlight(min(value) <= 10) + 
  labs(title = "Leaderboard over time - Ever in Top 10")

```

Even looking at those who finished in the top 10 shows that there is a lot 
of room to improve and even though you can have a big drop on the second day
there is still time to recover. 
```{r}
base_p + 
  gghighlight(last(value) <= 10) + 
  labs(title = "Leaderboard over time - Finished in Top 10")

```

Lastly let's take a look at anyone who was ever in the top two. There were
quite a few players tied for second on day three which must have made the last
round exciting since they were all within striking distance of the leader. 

```{r}
base_p + 
  gghighlight(min(value) <= 2, use_direct_label = FALSE) + 
  labs(title = "Leaderboard over time - Ever in second place") 


```

Choices for the pool were made based on groups of selections, three from group
A and B and two from group C. I assumed that these roughly translated to player
ability but it doesn't seem like group was an overly strong predictor of
performance. The cut includes players from all groups and the top performers in
the tournament were mostly those from group B. Generally it looks like the top performers in the pool were those that made strong picks from these groups since their size relative to group A also meant they were harder to pick from. 

Given the large amount of variability within groups I think my original idea
of fitting an HLM to identify players who performed better than their group
expectation would probably not be very interesting. The original idea there
was to look at predictions with/without shrinkage from hierarchical grouping
but I'm not convinced it will be worth the effort considering all groups have
good spread over all day four positions.

```{r fig.show="hold", out.width="50%", warning = FALSE}
score_pdat %>%
  ggplot(aes(x = name, y = value, group = player, colour = group)) +
  labs(x = "Day of Tournament", y = "Leaderboard Position") +
  scale_y_reverse() + 
  scale_colour_discrete(name = "Group") +
  geom_line() + 
  tx_def()

score_pdat %>%
  filter(name == "Day 4") %>%
  ggplot(aes(x = name, y = value, fill = group)) +
  labs(x = "Day of Tournament", y = "Leaderboard Position") +
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  scale_fill_discrete(name = "Group") +
  tx_def()
```


## Pool Performance

### Player Choices

Player choices in each group were degined by a smaller
number of "consensus" choices followed by a larger
spread. The winner of the pool tended to consistently
choose with the majority with the most start departure
coming in group A. 
```{r}

winner_picks <- entry_picks_anon %>% 
  filter(alt_name %in% "al-Saeed, Junaid") %>%
  select(-alt_name) %>% unlist

entry_picks_anon %>% select(-alt_name) %>%
  gather() %>%
  group_by(value) %>%
  summarise(picked = n()/nrow(entry_picks_anon)) %>%
  arrange(-picked) %>%
  mutate(value = forcats::as_factor(value),
         winner = ifelse(value %in% winner_picks, TRUE,FALSE),
         label = ifelse(winner == TRUE, round(picked,2), "")) %>%
  left_join(group_dict, by = c("value" = "player")) %>%
  mutate(group = paste("Group",group)) %>%
  ggplot(aes(x = value, y = picked, fill = winner)) +
  geom_col() +
  geom_text(aes(label = label), hjust = 0, size = 2, angle = 90) +
  tx_def() +
  labs(y = "Probability of Being Selected",
       x = "Player") +
  scale_fill_discrete(name = "Winner's Picks") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  facet_grid(. ~ group, scales = "free", space = "free_x")


```


Correlation plots with this many dimensions are not particularly interesting
or interpretable other than to say there are players that are selected together.
This is partly built into to the pool since they are built into groups. Given
that it might be interesting to look at plots separately.
```{r}
wide_dat %>% select(-alt_name) %>% cor(method = c("spearman")) %>%
  ggcorrplot(type = "lower",hc.order = TRUE,
             colors = c(plot_cols[[1]], "white", plot_cols[[2]])) +
  tx_def() +
  theme(axis.text.x = element_text(size = 3),
        axis.text.y = element_text(size = 3))


```

Looking at correlation plots by selection plot doesn't offer much extra insight
other than to provide some hope that there may indeed be some sort of pattern
in the golfers that are chosen together. We'll hope for a little a little more
insight into what's going on here when we get to PCA.

```{r}

g_dat <- group_dict %>% group_by(group) %>% nest

map2(g_dat$data, c(8, 6, 4), ~ {

  wide_dat %>% select(matches(.x %>% pull(player)))  %>% 
    cor(method = c("spearman")) %>%
  ggcorrplot(type = "lower",hc.order = TRUE,
             colors = c(plot_cols[[1]], "white", plot_cols[[2]])) +
  tx_def() +
  theme(axis.text.x = element_text(size = .y),
        axis.text.y = element_text(size = .y))
  
})


```



#### EVERSANA Choices

Let's take a high-level look at how those in the money chose compared to the
team from EVERSANA. It looks like you could get in the money with even a couple
poor performers or some exceptionally low ranking golfers. The most striking
difference between the top 5 and EVERSANA picks is that the EVERSANA picks are
almost uniformally distributed over most of the range of possible positions while
the top 5 generally had a high number of picks land in the top 5. Another trend
can be seen in where EVERSANA agreed (eg, Bryson Dechambeau) who was not selected
by any of the top 5 versus where they agreed (Justin Thomas) who was only selected
by one player from EVERSANA. In some respects this could just be a matter of seeing
that the people who won chose players with low scores and the people who lost didn't,
so we'll have to bring something a little more rigorous to this down the road.

```{r}

top_5 <- final_scores %>% 
  arrange(score_4) %>% 
  slice(1:5) %>% 
  pull(alt_name)

entry_picks_anon %>% 
  filter(alt_name %in% c(names(test_names), top_5)) %>%
  pivot_longer(-alt_name) %>%
  left_join(score_pdat %>% filter(name == "Day 4") %>%
  select(player, value) %>%
    rename(rank = value,
           value = player)) %>%
  arrange(rank) %>%
    left_join(final_scores %>% select(alt_name, score_4)) %>%
  arrange(score_4) %>%
  mutate(
    winner = ifelse(alt_name %in% top_5, "Top 5", "EVERSANA"),
    alt_name = glue::glue("{alt_name} ({score_4})"),
    value = glue::glue("{value} ({rank})"),
    alt_name = forcats::as_factor(alt_name)
    ) %>%
  arrange(rank) %>%
  mutate(value = forcats::as_factor(value)) %>%
  ggplot(aes(x = alt_name, y = value, colour = winner)) +
  geom_point() +
  tx_def() +
  scale_colour_es(name = "Top 5 vs EVERSANA", "primary") +
  labs(title = "EVERSANA Player Selections", x = "", y = "") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  


```

## Player Performance

Lastly let's take a look at player performance. We saw that with golfers there
was a big swing in the ranks over time, but how did that translate into the pool?

### Overall

```{r}
p_bp <- 
  final_scores %>%
  gather("key", "value", -c(alt_name)) %>%
  mutate(key = factor(key, labels = c("Day 1", "Day 2", "Day 3", "Day 4"))) %>%
  group_by(key) %>%
  arrange(key, value) %>%
  mutate(pos = 1:n()) %>%
  group_by(key, value) %>%
  mutate(pos = min(pos)) %>%
  group_by(alt_name) %>%
  mutate(sd = sd(pos),
         diff = max(pos) - min(pos)) %>%
  ggplot(aes(x = key, y = pos, group = alt_name)) +
  labs(x = "Day of Tournament", y = "Total Score") +
  scale_y_reverse() + 
  geom_line() + 
  tx_def()
 
p_bp + labs(title = "Rank - All players")
  

```

So similar to the golfer situation we can see a lot of people were in the top ten
at some point and there are some pretty drastic flips. We can see maybe a group
of people who started of well and stayed doing well, an opposite group, and then
a bunch of zig-zags somewhere in the middle. 
```{r}
p_bp + 
  gghighlight(min(pos) <= 10) + 
  labs(title = "Leaderboard over time - Ever in Top 10") +
  geom_hline(yintercept = 10, colour = plot_cols[[2]])
```

What's prety interesting is that it seems that most people who finished in the top
10 didn't start there and almost half were out of the top 10 on day three. That said
the winner more or less declared themselves on day two and never went back. 
```{r}

p_bp + 
  gghighlight(last(pos) <= 10) + 
  labs(title = "Leaderboard over time - Finished in top 10") +
  geom_hline(yintercept = 10, colour = plot_cols[[2]])


```

Now let's do the opposite of this and take a look at the worst performers. It
looks like if you had a bad first day you could escape doing very poorly but
it was difficult to ever to better than 150 if you were ever worse than 200. We can
also see that starting well doesn't mean anything, presumably because you a bad
second day across multiple players is impossible to recover from. 

```{r}
p_bp + 
  gghighlight(max(pos) > 200) + 
  labs(title = "Leaderboard over time - Ever worse than 200") +
  geom_hline(yintercept = 200, colour = plot_cols[[2]])
```

I think one of the most fascinating things about this tournament rank data is 
how much variability people have across days. So let's highlight some groups
with high variability. Using a cut-point of 50 for the standard deviation in
position provides a nice grop of players who had an exciting pool. This includes
people who started high and had precipitous drops as well as the opposite and then
people who are somewhere in-between.

```{r}

p_bp + 
  gghighlight(sd > 50) + 
  labs(title = "Leaderboard over time - Rank SD > 50") +
  geom_hline(yintercept = 200, colour = plot_cols[[2]])

```

Another way of looking at this is to find players who had a difference of more
than 150 in their best and worst scores. 
```{r}

p_bp + 
  gghighlight(diff > 150) + 
  labs(title = "Leaderboard over time - Difference in best and worst rank > 150") +
  geom_hline(yintercept = 200, colour = plot_cols[[2]])

```

Finally let's see the EVERSANA crew.

```{r}

p_bp + 
  gghighlight(alt_name %in% names(test_names)) + 
  labs(title = "Leaderboard over time - Difference in best and worst rank > 150") +
  geom_hline(yintercept = 200, colour = plot_cols[[2]])

```
