---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Initial Data Exploration


## Golfer Performance

If we look at the leaderboard position of the full player field over time
we end up with something that is a little overwhelming. We can several ties
over time where ranks converge together and we can also see the cuts on day two 
from the straight lines from that point onwards. One thing I am a little surprised by in the beginning is that once you get out of the best few players
you actually get a lot of variability in ranks over days. This makes me regret
not taking a little more time to watch because I can see already that daily
scores in the pool would have had some potentially big swings.

```{r}
# Match style being used in the slides

rank_names <- c(r1_rank = "Day 1",
                r2_rank = "Day 2",
                r3_rank = "Day 3",
                r4_rank = "Day 4")

# Add indicator for players that are cut
score_dat <- score_dat %>%
  mutate(cut = ifelse(is.na(pos), TRUE, FALSE))

tx_def <-  function() {theme_xaringan(text_font_size = 10,
                 title_font_size = 10)
}


score_pdat <- score_dat %>%
  select(player, r1_rank:r4_rank, cut) %>%
  pivot_longer(-c(player, cut)) %>%
  mutate(name = factor(name, levels = names(rank_names), labels = rank_names)) %>%
  left_join(group_dict)

base_p <- score_pdat %>%
  ggplot(aes(x = name, y = value, group = player)) +
  labs(x = "Day of Tournament", y = "Leaderboard Position") +
  scale_y_reverse() + 
  geom_line() + 
  tx_def()
 
base_p + labs(title = "Leaderboard over time - All players")
  

```

I don't really watch a lot of golf so coming in I was expecting players that
were cut were generally would have had a bad tournament from the beginning.
But we can see that while this is broadly true there were actually a few players
who came out strong but had a bad day two. 

```{r}
base_p +
  gghighlight(cut == TRUE)
```

We can look a little closer at the development of the top ten players over time.
It looks like you get a mix of people who pretty much started and stayed there,
and others who had a bad first day before finishing strong. 
```{r}
base_p + 
  gghighlight(min(value) <= 10) + 
  labs(title = "Leaderboard over time - Ever in Top 10")

```

Even looking at those who finished in the top 10 shows that there is a lot 
of room to improve and even though you can have a big drop on the second day
there is still time to recover. 
```{r}
base_p + 
  gghighlight(last(value) <= 10) + 
  labs(title = "Leaderboard over time - Finished in Top 10")

```

Lastly let's take a look at anyone who was ever in the top two. There were
quite a few players tied for second on day three which must have made the last
round exciting since they were all within striking distance of the leader. 

```{r}
base_p + 
  gghighlight(min(value) <= 2, use_direct_label = FALSE) + 
  labs(title = "Leaderboard over time - Ever in second place") 


```

Choices for the pool were made based on groups of selections, three from group
A and B and two from group C. I assumed that these roughly translated to player
ability but it doesn't seem like group was an overly strong predictor of
performance. The cut includes players from all groups and the top performers in
the tournament were mostly those from group B. Generally it looks like the top performers in the pool were those that made strong picks from these groups since their size relative to group A also meant they were harder to pick from. 

Given the large amount of variability within groups I think my original idea
of fitting an HLM to identify players who performed better than their group
expectation would probably not be very interesting. The original idea there
was to look at predictions with/without shrinkage from hierarchical grouping
but I'm not convinced it will be worth the effort considering all groups have
good spread over all day four positions.

```{r fig.show="hold", out.width="50%", warning = FALSE}
score_pdat %>%
  ggplot(aes(x = name, y = value, group = player, colour = group)) +
  labs(x = "Day of Tournament", y = "Leaderboard Position") +
  scale_y_reverse() + 
  scale_colour_discrete(name = "Group") +
  geom_line() + 
  tx_def()

score_pdat %>%
  filter(name == "Day 4") %>%
  ggplot(aes(x = name, y = value, fill = group)) +
  labs(x = "Day of Tournament", y = "Leaderboard Position") +
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  scale_fill_discrete(name = "Group") +
  tx_def()
```



## Pool Performance

### Player Choices

```{r}
entry_picks_anon
```

## Player Performance

## Overall

## Team Highlight

```{r}
final_scores

```
