---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Results Analysis

## Player Choice Clustering


### Example

A brief set of example code was provided by Ryan based on [this tutorial](https://www.datacamp.com/community/tutorials/pca-analysis-r
). The analysis is based on the mtcars dataset and provides a very bare bones
showcase of the `prcomp` function and `ggbiplot`.

First we create the PCA object using `prcomp` limited to continuous variables only.
This was done because the other variables in mtcars are categorical/ordinal which
do not work well with PCA without further processing.

```{r}
mtcars.pca <- prcomp(mtcars[,c(1:7,10,11)], center = TRUE, scale. = TRUE)

summary(mtcars.pca)

```

We can them make a simple plot to show how different variables are related. The
post goes into some extra details around drawing ellipses etc but we won't get
into that until the analysis phase if there is something interesting. One
of the concepts that they showed that was pretty interesting though was how this
type of plot can help you to identify variables that might be important for
classification problems. For example all of the american cars cluster tightly
in one section of the plot, suggesting they would be important in an "American vs
not American" classification. 

```{r fig.width= 8, fig.height= 8}
ggbiplot(mtcars.pca,ellipse=TRUE,choices=c(3,4),   labels=rownames(mtcars))

```



## Examples from Ryan

Ryan made a nice start on some basic line plots that we can build on
from here.
```{r}
golf_data <- tribble(
  ~name, ~day, ~pos, ~icon,
  "T", "1", 104, NA,
  "C", "1", 102, NA,
  "I", "1", 145, NA,
  "J", "1", 164, NA,
  "R", "1", 170, NA,
  # Day Two
  "T", "2", 67, "/home/rbatten/learning-r/masterChirpsRBcalcs/masters-jacket.png",
  "C", "2", 121, NA,
  "I", "2", 200, NA,
  "J", "2", 109, NA,
  "R", "2", 117, NA
)

ggplot(golf_data,
       mapping = aes(x = day, y = pos,
                                group = name)) +
  scale_y_reverse() +
  geom_line(aes(color = name)) +
  geom_point() +
  theme_minimal() +
  ggtitle("Master's Pool Results") +
  scale_color_manual(values = c("T" = "#006B54", "C" = "#DC4605",
                                "I" = "#007298", "J" = "#ED8B00",
                                "R" = "#002F6C")) +
  geom_image(aes(image = icon), size = 0.05)

```




